---
title: "Salaries in Seattle District"
author: "bh"
date: "May 17, 2019"
output:
  pdf_document:
    toc: yes
  html_document:
    df_print: paged
    toc: yes
  word_document:
    toc: yes
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=4, fig.height=3,
                      echo=FALSE, warning=FALSE, message=FALSE)
```

#big note: locate where ptf is grepl'd because it will not work for most years


#Intro
This is a quick set of plots about salary based on some very rough filtering by job title.  

##Used packages in R
```{r load packages,echo=T,results='hide'}
library(tidyverse)
library(dplyr)
library(lubridate)
library(scales)
library(tidytext)
library(stringr)
```


#Code used to wrangle or tidy

This data comes from the washington state salary database.

```{r load data, include =F, results='hide'}
df <- read.csv(file="http://facweb.northseattle.edu/bholt/quickshare/AnnualEmployeeSalary2013thru2017.csv",header=T)

```
```{r basic tidying, echo=T}
(var.name<- names(df))
var.name[5:9] <- c('2013','2014','2015','2016','2017')
names(df) <- var.name
#the following is an HR position that is captured by my later grep of faculty
df$JobTitle <- sub("SPEC, FACULTY DEV.& COMPLI","SPEC, HR DEV.& COMPLI",df$JobTitle,ignore.case=T) 
df$JobTitle <- trimws(df$JobTitle) #takes out unnecessary whitespace
df$Name <- trimws(df$Name) #takes out unnecessary whitespace
df$AgyTitle <- trimws(df$AgyTitle) #takes out unnecessary whitespace
```
##This locates the agency code
```{r find district agency number,echo=T}
df %>% filter(
  grepl("seattle",AgyTitle,ignore.case = T)) %>% 
  select(Agy) %>% head(2)

df %>% filter(Agy==670,grepl("faculty|ftf",JobTitle,ignore.case = T)) %>%  
    summarise(n()) #just a check

df %>% 
  filter(grepl("seattle",
 AgyTitle,ignore.case = T)) %>% select(Agy) %>% head(2) #another check

```

##Selecting just Seattle data, followed by tidying (campus data unavailable)
```{r select seattle and tidy the dataframe, echo=T,results='hide'}
gath.df<- df %>%  #creates seattle data in long format
  filter(Agy==670) %>% 
  gather(key=year,value=sal,-c(Agy:JobTitle)) %>% filter(sal>0)

head(gath.df)

#change date format using lubridate's ymd (should do this earlier)
gath.df <- mutate(gath.df,year=ymd(paste0(gath.df$year,"06-30-")))

#check
gath.df %>% group_by(year) %>%   summarise(mean(sal))

# very rough filter  for faculty
FTF.grepl <- "faculty|ftf"
PTF.grepl <- 'part|ptf'   ##big note here.  PTF is only recorded since 2016
gath.df %>% filter(grepl(FTF.grepl,JobTitle,ignore.case = T)) %>%
  group_by(year) %>% summarise(n(),mean(sal),min(sal),max(sal))
```

```{r tidytext, echo=T}
#titles<- tibble(lines=1:length(gath.df$JobTitle),text=gath.df$JobTitle)
tidy.titles<- titles %>% unnest_tokens(word,text)
tidy.titles <- tidy.titles %>% anti_join(stop_words)
tidy.titles %>% count(word,sort=T) %>% filter(grepl('chance',word,ignore.case = T))

gath.df <- gath.df %>% mutate(year_group=year(year),line=row_number())

tidy_jobtitles<- gath.df %>% unnest_tokens(word,JobTitle)
data(stop_words)

tidy_jobtitles <- tidy.jobtitles %>%
  anti_join(stop_words)
clus.df<- tidy_jobtitles %>% group_by(year_group) %>% 
  count(word, sort = T) %>% filter(n<5)

```
```{r cluster titles,echo=T}
clus.df

```

#Mean of salary
```{r plot means of sal, echo=T}
dean.grep <- 'dean|director|^dir|dir'

plotting.df<- gath.df %>% 
  mutate(yr=ifelse(year=="2013-06-30",'2013',
              ifelse(year=="2014-06-30",'2014',
              ifelse(year=="2015-06-30",'2015',
              ifelse(year=="2016-06-30",'2016','2017'))))) %>% 
  mutate(status=ifelse(
                  grepl(FTF.grepl,JobTitle,ignore.case =T)==T,'FTF',
                ifelse(grepl(PTF.grepl,JobTitle,
                 ignore.case =T)==T,'PTF',
                 ifelse(grepl(dean.grep,JobTitle,
                 ignore.case =T)==T,'dean/dir','non-fac')))) %>%
  group_by(yr,status) %>% 
  summarise(n=n(),m=mean(sal),sd=sd(sal))

head(plotting.df)
```

```{r plot mean, echo=T}
plotting.df %>% 
  ggplot(aes(x=yr,y=m,group=status,color=status))+
  geom_point()+
  geom_line()+ 
  scale_y_continuous(labels = scales::comma)+
  ggtitle("Seattle College averages in Salary") +
  xlab("Year") + ylab("Mean Salary in Dollars")
```



##Plot salaries by quantiles
```{r plot mean quantiles, echo=T, fig.height=4, fig.width=6}

gath.df %>% 
  mutate(yr=ifelse(year=="2013-06-30",'2013',
              ifelse(year=="2014-06-30",'2014',
              ifelse(year=="2015-06-30",'2015',
              ifelse(year=="2016-06-30",'2016','2017'))))) %>% 
  mutate(status=ifelse(
                  grepl(FTF.grepl,JobTitle,ignore.case =T)==T,'faculty',
                ifelse(grepl(dean.grep,JobTitle,
                 ignore.case =T)==T,'dean/dir','non-fac'))) %>%
  group_by(yr,status) %>% filter(sal>14000) %>% 
  summarise(n=n(),
          Salary=list(enframe(quantile(sal, probs=c(0,0.25,0.5,0.75,1))))) %>% 
  unnest -> plot.data

  plot.data %>% 
  ggplot(aes(x=yr,y=value,group=status,color=status))+
  geom_point()+ggtitle("Seattle College Salary Qantiles") +
  xlab("Year") + ylab("Mean Salary in Dollars")+
  scale_y_continuous(labels = scales::comma)+
  facet_wrap(~status)+
  geom_line(aes(x=yr,y=value,group=name),linetype='dashed')+
  geom_smooth(aes(x=yr,y=value,group=name),method = 'lm',size=.1,se=F)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))


plot.data %>% 
  ggplot(aes(x=yr,y=n,group=status,color=status))+geom_point()+
  geom_line()+facet_wrap(~status)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+ggtitle('number of employees by status')

```

```{r sum of status over time, echo=T}
gath.df %>% 
  mutate(yr=ifelse(year=="2013-06-30",'2013',
              ifelse(year=="2014-06-30",'2014',
              ifelse(year=="2015-06-30",'2015',
              ifelse(year=="2016-06-30",'2016','2017'))))) %>% 
  mutate(status=ifelse(
                  grepl(FTF.grepl,JobTitle,ignore.case =T)==T,'faculty',
                ifelse(grepl('^dir|dean|director',JobTitle,
                 ignore.case =T)==T,'dean/dir','non-fac'))) %>%
  group_by(yr,status) %>% 
  summarise(n=n(),sum.of.Sal=sum(sal)) %>%
  
  ggplot(aes(x=yr,y=sum.of.Sal,color=status,group=status))+
  geom_line()+
  scale_y_continuous(labels = scales::comma)+theme(axis.text.x = element_text(angle = 60, hjust = 1))+ggtitle('sum of all salary\nby status' )

```

#Differences in Salary across years

```{r difference in total workers across years,echo=T}
gath.df %>% 
  mutate(yr=ifelse(year=="2013-06-30",'2013',
                   ifelse(year=="2014-06-30",'2014',
                          ifelse(year=="2015-06-30",'2015',
                                 ifelse(year=="2016-06-30",'2016','2017'))))) %>% 
  mutate(status=ifelse(
    grepl(FTF.grepl,JobTitle,ignore.case =T)==T,'faculty',
    ifelse(grepl(dean.grep,JobTitle,
                 ignore.case =T)==T,'dean/dir','non-fac'))) %>% 
  group_by(yr) %>% 
  summarise(sum.sal = sum(sal)) %>% 
  #group_by(yr) %>% 
  summarise(sum.dif=list(diff(sum.sal))) %>% unnest(sum.dif)->sum.dif 

cbind(as.data.frame(sum.dif),data.frame(yrs=c(2014,2015,2016,2017))) %>% 
  ggplot(aes(x=yrs,y=sum.dif))+geom_line()+
  scale_y_continuous(labels = scales::comma)+
  ggtitle("Difference in spending from previous year") +
  xlab("Year") + ylab("Difference in Salary ")

```

#differences in spending on salary by group
```{r difference in grouped workers across years,echo=T}
gath.df %>% 
  mutate(yr=ifelse(year=="2013-06-30",'2013',
                   ifelse(year=="2014-06-30",'2014',
                          ifelse(year=="2015-06-30",'2015',
                                 ifelse(year=="2016-06-30",'2016','2017'))))) %>% 
  mutate(status=ifelse(
    grepl(FTF.grepl,JobTitle,ignore.case =T)==T,'faculty',
    ifelse(grepl(dean.grep,JobTitle,
                 ignore.case =T)==T,'dean/dir','non-fac'))) %>% 
  group_by(status,yr) %>% 
  summarise(n=n(),sum.sal = sum(sal) )-> sum.Salaries.by.grp.yr
  
 df.sum.dif<- data.frame( cbind(data.frame(sum.Salaries.by.grp.yr),data.frame(sum.dif=c(0,diff(sum.Salaries.by.grp.yr$sum.sal)))))

 for(i in 1:nrow(df.sum.dif)){ 
   if(df.sum.dif$yr[i]==2013){df.sum.dif$sum.dif[i]=0 }
 }
 
       df.sum.dif %>% filter(yr!=2013) %>% 
 ggplot(aes(x=yr,y=sum.dif,group=status,color=status))+geom_line()+
  scale_y_continuous(labels = scales::comma)+
  ggtitle("Difference in spending from previous year") +
   xlab("Year") + ylab("Difference in Summed Salary ")+facet_wrap(~status)

```

#Standard Deviation of salary
##(it's not really helpful)
I thought I'd look at this but because I'm grouping by job title, in some cases there are only single values so there isn't any standard deviation to be calculated in a given year.
```{r plot standard deviation, echo=T}
plotting.df %>% 
  ggplot(aes(x=yr,y=sd,group=status,color=status))+
  geom_point()+
  geom_line()+ 
  scale_y_continuous(labels = scales::comma)+
  ggtitle("Seattle College standard deviation in Salary") +
  xlab("Year") + ylab("Standard Deviation of Sal in Dollars")
```


```{r 2016 dean variation,include=F}

gath.df %>% 
  mutate(yr=ifelse(year=="2013-06-30",'2013',
              ifelse(year=="2014-06-30",'2014',
              ifelse(year=="2015-06-30",'2015',
              ifelse(year=="2016-06-30",'2016','2017'))))) %>% 
  mutate(status=ifelse(
                  grepl(FTF.grepl,JobTitle,ignore.case =T)==T,'faculty',
                ifelse(grepl('dean|director',JobTitle,
                 ignore.case =T)==T,'dean/dir','non-fac'))) %>%
  group_by(yr,JobTitle) %>% 
  filter(status=='dean/dir',yr=='2016') %>% summarise(n(),mean(sal),sd=sd(sal))
```
#Future

What would be nice is a quick way to get salary categories so we could show differences between staff (and their subgroups) compared to exempt (Vice pres/chancellors, etc).  
